---
title: "Predictive Barbell Lifts"
author: "Ted Powers"
date: "Jan 21, 2015"
output: html_document
---

###Summary
In this project, data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants will be used.  From this data, the goal is to predict which of the 5 classes of dumbell lifts was performed and whether the exercise was performed correctly.  

In order to reproduce the same results, you need a certain set of packages, as well as setting a pseudo random seed equal to the one I used.

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
library(klaR)
library(abind)
library(arm)
library(caret)
#library(ggplot2)
library(randomForest)
library(kernlab)

library(rattle)

set.seed(42)
```


###Data
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 


```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
setwd("~/GitHub/machinelearning1")
if (!dir.exists("data")) { dir.create("data")}

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainFile <- "train.csv"
trainFilePath <- paste(getwd(),"data", trainFile, sep = "/")

testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testFile <- "test.csv"
testFilePath <- paste(getwd(),"data", testFile, sep = "/")

## Download Files from internet only if not in data directory
if (!file.exists(paste(getwd(), "data", "train.csv", sep ="/"))) {
        setInternet2(use = TRUE)
        download.file(trainUrl, destfile = trainFilePath)
}

if (!file.exists(paste(getwd(), "data", "test.csv", sep ="/"))) {
        setInternet2(use = TRUE)
        download.file(testUrl, destfile = testFilePath)
}

## load the file from disk only if it 
## hasn't already been read into a variable
if(!(exists("tran"))) {
        tran <- read.csv(trainFilePath, na.strings = c("NA","#DIV/0!",""))
}
if(!(exists("test"))) {
        test <- read.csv(testFilePath, na.strings = c("NA","#DIV/0!",""))
}

rm(testFile, testFilePath, testUrl, trainFile, trainFilePath, trainUrl)
```

###Training-Set cleaning and pre-processing

Names’s Coherence Check
```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
all.equal(colnames(test)[1:length(colnames(test))-1], colnames(tran)[1:length(colnames(tran))-1])
```

To ease the computation and due to the low informativity loss, the dataset is cleaned from the variables with an high share of NAs and from the ones characterized by low variance.

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
nearzero <- nearZeroVar(tran, saveMetrics = TRUE)

tran_orig <- tran

tran <- tran[, !nearzero$nzv]

```

Variables with more than 20% missing values are removed

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
toberem <- sapply(colnames(tran), function(x) if(sum(is.na(tran[, x])) > 0.20*nrow(tran))
        {return(TRUE)}
        else{return(FALSE)
}
)
tran_t1 <- tran

tran <- tran[, !toberem]
```

Variables related with data acquisition (like: id, timestamps, individuals’ names, etc.) are not suitable to be used in prediction and are removed
```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
tran_t2 <- tran
tran <- tran[, -(1:6)]
```

Correlation analysis:
```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
Hcorr <- caret::findCorrelation(cor(tran[, -53]), cutoff=0.6)
names(tran)[Hcorr]
```

Many variables are highly correlated. PCA will be used in the pre-processing. After the data cleaning the variables selected to specify the model are:
```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
names(tran)
```


###Partioning the training set into two
Partioning Training data set into two data sets, 60% for myTrain, 40% for myTest:

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
inTrain <- createDataPartition(y=tran$classe, p=.6, list=FALSE)

tran_t3 <- tran
myTest <- tran[-inTrain, ]
tran <- tran[inTrain, ]
dim(tran); dim(myTest)

```

```{r, echo=T, message=FALSE, warning=FALSE, results="as-is"}
head(test)
head(tran)

```

##Model Specification and Cross Validation

In order to avoid overfitting and to reduce out of sample errors, TrainControl is used to perform 7-fold cross validation.

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
tc <- trainControl(method = "cv", number = 7, verboseIter=FALSE , preProcOptions="pca", allowParallel=TRUE)
```

Six models are estimated: Random forest, Support Vector Machine (both radial and linear), a Neural net, a Bayes Generalized linear model and a Logit Boosted model.

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
rf <- train(classe ~ ., data = tran, method = "rf", trControl= tc)
svmr <- train(classe ~ ., data = tran, method = "svmRadial", trControl= tc)
NN <- train(classe ~ ., data = tran, method = "nnet", trControl= tc, verbose=FALSE)
svml <- train(classe ~ ., data = tran, method = "svmLinear", trControl= tc)
bayesglm <- train(classe ~ ., data = tran, method = "bayesglm", trControl= tc)
logitboost <- train(classe ~ ., data = tran, method = "LogitBoost", trControl= tc)
```

Accuracy comparision

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
model <- c("Random Forest", "SVM (radial)","LogitBoost","SVM (linear)","Neural Net", "Bayes GLM")
Accuracy <- c(max(rf$results$Accuracy),
        max(svmr$results$Accuracy),
        max(logitboost$results$Accuracy),
        max(svml$results$Accuracy),
        max(NN$results$Accuracy),
        max(bayesglm$results$Accuracy))
        
Kappa <- c(max(rf$results$Kappa),
        max(svmr$results$Kappa),
        max(logitboost$results$Kappa),
        max(svml$results$Kappa),
        max(NN$results$Kappa),
        max(bayesglm$results$Kappa))  

performance <- cbind(model,Accuracy,Kappa)
knitr::kable(performance)
```

Random forest and SVM(radial) provide the best results and will provide the predictions for the submission. Even if the Out of sample error cannot be estimated exactly, the in-sample error obtained through cross-validation is calculated over different test sets and should provide a better estimate of out-of sample error with respect to the case of no cross-validation.

##Prediction of “classe” variable for the test set

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
rfPred <- predict(rf, test)
svmrPred <- predict(svmr, test)
```

Checking if the models give same predictions

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
prediction <- data.frame(cbind(rfPred, svmrPred))
prediction$same <- with(prediction, rfPred == svmrPred)
colnames(prediction) <- c("Random Forest", "SVM (radial)", "Same Prediction")
knitr::kable(prediction)
```

Generation of the files to be submitted is made through the provided function

```{r, echo=F, message=FALSE, warning=FALSE, cache = TRUE, results="hide"}
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}

pml_write_files(rfPred)
pml_write_files(svmrPred)
```

##Conclusions
The random forest model provides an outstanding accuracy and, accordingly, the predictions for the test set were correct in 100% of the cases.
